---
title: "Modeling_transmission"
output: html_notebook
---

To predict pairwise values with other pairwise values you can take two routes: Null-model approach (matrix permutation models) or dyadic regression with multimembership random effect structure. The former is using the dependence structure of the data itself as a null model so you won't need to know much about it, just see whether the correlation between two matrices is more extreme than the distribution of correlation coefficients of the same two matrices when the other one is randomized multiple times and correlation re-calculated each time. The second is a linear model approach, that aims at accounting for the specific type of dependence introduced by pairwise comparisons with multimembership random effect structure. This approach is more complicated but also more flexible in that it can accommodate a variety of different response variable types (proportions, counts etc...) with different distributions (skewed, normal), it can deal with interaction effects and repeated measurements/samples per individual (nested autocorrelation) alongside the dyadic dependence of the values (multimembership autocorrelation).

As and example, we will use a real data set on wild wood mouse microbiome with known correlation patterns between dyadic variables.  We will construct a social network based on spatio-temporal co-occurrence (how often two individuals are observed in the same place within a short time window) and ask whether social network association strength predicts microbiota sharing (similarity) between two mice, independent of their spatial distance, temporal distance and other categorical covariates (e.g. whether individuals are of the same or different age and sex).

Here, the variables of interest are:
- Microbiota similarity (Jaccard index of similarity = proportion of shared taxa of all taxa of a pair)
- Sex similarity (binary= same or different)
- Spatial distance (meters between location centroids of individual mice, based on tracking data (="loggerdata")
- Temporal distance (days between collection between any two microbiome samples)
- Social association strength (Proximity/ edge weight in the social network.

All variables will be scaled between 0-1 for modeling to enhance comparability of effect estimates.

This code will walk through the process in order or:

1. Making the pairwise model data from real-life data set 
  
2. Constructing the MRQAP matrix permutation model using asnipe package

3. Constructing the dyadic Bayesian regression model using MCMCglmm 

-----------------------------------------------------------------------------

1. Making pairwise data:
Making dyadic data matrices from the raw data. Here I will make four types of pairwise matrices describing pairwise variables between the same set of individuals:

1.1 continuous similarity matrix (here, microbiome Jaccard similarity matrix)
1.2 continuous distance matrices (here, spatial distance matrix and temporal distance matrix)
1.3 continuous association matrix (here, social network based on spatiotemporal co-occurrence)
1.4 binary similarity matrices (here, sex similarity), 

I will then show hoe to unravel these matrices to create a dyadic data frame
1.5. Dyadic data frame


-----------------------------------------------------------------------------
1.1 Making microbiome similarity matrices to describe ASV sharing (here,Jaccard similarity):
```{r}
library(phyloseq)

#Read in microbiome data and associated sample data in phyloseq format. These data sets contain either 70 samples of 70 individuals
phy<-readRDS( "Mouse_microbiome_idwise.rds")

#Sample data inside this phyloseq object contains information of each microbiome samples, such as which individuals they came from, what sex these individuals were and when the sample was collected:
sample_data<-sample_data(phy)
sample_data[,c("ID","Sample_name","Sex", "date")]

#make key for the order of sample names and associated individuals.
key<-data.frame(ID=sample_data(phy)$ID, Sample_name=sample_data(phy)$Sample_name)

#Make Jaccard dissimilarity matrix from microbiome data using vegdist function embedded in phyloseq::distance
JACM<- as.matrix(phyloseq::distance(phy, method = "jaccard", type = "samples"))
#Transpose Jaccard dissimilarity marix to Jaccard similarity matrix
JACM<-1-JACM

#Note that Jaccard similarity matrix has rownames and colnames in the same order as key
all(rownames(JACM)==key$Sample_name)
# So we can just call the rownames and colnames with the names of individuals since there is just one sample per individual in this data set
rownames(JACM)<-key$ID
colnames(JACM)<-key$ID

#See Jaccard similarities across individuals
JACM

#Save matrix to ready matrices folder
saveRDS(JACM,"ready_matrices/JACM.rds")
```
1.2 Making continuous spatial and temporal distance matrices.

A) Temporal distance matrix 
this matrix will describe the distance in days between microbiome samples based on the dates recorded in sample_data 
```{r}
#Transform dates into a numeric variable
sample_data$date<-as.Date(as.character(sample_data$date), format="%d/%m/%Y")
sample_data$date_numeric<-as.numeric(sample_data$date)

#Create data frame with each sample name (character) and sampling time (numeric) 
SampleTime_frame<-sample_data[,c("Sample_name","date_numeric")]
SampleTime_frame$Sample_name<-as.character(SampleTime_frame$Sample_name)

#Create an empty matrix to fill with distances
TEMPM<-array(0,c(nrow(SampleTime_frame),nrow(SampleTime_frame)))

#Derive matrix with time distances between each sample
for (i in 1:nrow(SampleTime_frame)){
 for (j in 1:nrow(SampleTime_frame)) 
{TEMPM[i,j]=abs(SampleTime_frame$date_numeric[i] -SampleTime_frame$date_numeric[j])
  }
}

#Note that Temporal distance matrix has rownames and colnames in the same order as key
all(rownames(TEMPM)==key$Sample_name)
# So we can just call the rownames and colnames with the names of individuals since there is just one sample per individual in this data set

#Name rown amd colnames with individual names 
rownames(TEMPM)<-key$ID
colnames(TEMPM)<-key$ID

#Save matrix to ready matrices folder
saveRDS(TEMPM,"ready_matrices/TEMPM.rds")
```
B) Spatial distance matrix 
this matrix will describe the distance in meters between the location centroid of each individual mouse, based on loggerdata

Loggerdata is a data set where each row is a time-stamped observation of an individual mouse (ID) in a specific location (described with coordinates x and y). This data can be used to derive social networks as well as spatial centroid matrices of mice.

```{r}
Loggerdata<- readRDS("Loggerdata_simple.rds")
Loggerdata_reduced<-Loggerdata[,c("ID","X_coord","Y_coord")]

#Find mean location coordinates (centroid of logger records) for each individual
centroids<-aggregate(Loggerdata_reduced[,2:3], by=list(Loggerdata_reduced$ID),FUN=mean,na.rm=TRUE)
colnames(centroids)<-c("ID","X_coord_mean","Y_coord_mean")

#Create an empty matrix to fill with distances
SPATM<-array(NA,c(length(centroids$ID),length(centroids$ID)))

#Derive matrix with spatial distances between each location centroid
for (i in 1:length(centroids$ID)){
 for (j in 1:length(centroids$ID)) 
{SPATM[i,j]= sqrt((abs(centroids$X_coord_mean[i]-centroids$X_coord_mean[j]))^2+(abs(centroids$Y_coord_mean[i]-centroids$Y_coord_mean[j]))^2)
  }
}

#Note that Association matrix has rownames and colnames in the same order as key
all(rownames(SPATM)==key$ID)

#Save matrix to ready matrices folder
saveRDS(SPATM,"ready_matrices/SPATM.rds")
```
1.3 Making a social network based on spatiotemporal co-occurrence in the loggerdata using standard functions of the R package asnipe.

This uses the night-location combination as the spatio-temporal grouping (individuals observed more often in the same location during same night are considered more "associated") 
and calculates social association through "Simple Ratio Index" (SRI) as follows:

SRI= X/X+ya+yb+yab,

where X is the number of instances (nights) where individual a and b were observed "associated" (=in the same location during that night), ya is the number of instances where individual a was observed but b was not, yb is the number of instances where individual b was observed but a was not and yab is the number of instances where individuals a and b were both observed but not associated (not observed in the same location that night).

This is probably the crudest form of social association measure I know, and there are also much more delicate ones.

```{r}
library(asnipe)

# Make a spatio-temporal grouping variable by combining variables "logger" (=unique location) and "night" (=unique time)

Loggerdata$lognight_logger<-paste(Loggerdata$night, Loggerdata$logger, sep="-")

#Aggregate Loggerdata per individual
t<-table(Loggerdata$ID, Loggerdata$lognight_logger)
log<-t>0

#Make "group-by-individual matrix" (gbi), where lognight-logger is considered the grouping variable
gbi<- replace(t, log, 1)
gbi<-t(gbi) # Here individuals as columns and lognights as rows 

#derive social association matrix using the default Simple Ratio Index method of asnipe package 
AM<-get_network(gbi, association_index="SRI")

#Note that Association matrix has rownames and colnames in a different order from key
all(rownames(AM)==key$ID)
# We need to re-order the matrix to be in the same order with the other matrices

AM<-AM[match(key$ID,rownames(AM)), match(key$ID,colnames(AM))]

#Now Association matrix has rownames and colnames in same order as key
all(rownames(AM)==key$ID)

#Save matrix to ready matrices folder
saveRDS(AM,"ready_matrices/AM.rds")

```
Fun EXTRA: plotting social network with igraph package
```{r}
library(igraph)
#Create an adjacency matrix
net<- graph.adjacency(AM, mode= 'undirected', weighted=TRUE, diag=FALSE)

#metadata for the plot
demo<-sample_data[,c("ID","Sex")]
demo<-unique(demo)
demo<-demo[match(rownames(AM),demo$ID),]
demo$weighted_degree<- strength(net, vids=V(net))

sexes<-unique(demo$Sex)
sexes.numeric<- as.numeric(demo$Sex)
demo$sexcolor<-rainbow(length(sexes))[sexes.numeric]
demo$sexcolor<-as.factor(demo$sexcolor)
levels(demo$sexcolor)<-c("red","blue")
demo$sexcolor<-as.character(demo$sexcolor)

plot(net, vertex.color=demo$sexcolor, vertex.label=NA,vertex.frame.color = demo$sexcolor, vertex.size=demo$weighted_degree^0.8,  edge.color="black",edge.width= E(net)$weight*5, edge.curved=0.4)
```
1.4 Making binary similarity matrices (here, sex similarity), based on individual info stored in sample_data. The resulting matrix will have for each individual pair a value of "1"= same sex or "2"=different sex
```{r}

#Create data frame with each Individual name (character) and their sex (Character)
Sex_frame<-sample_data[,c("ID","Sex")]
Sex_frame$ID<-as.character(Sex_frame$ID)
Sex_frame$Sex<-as.character(Sex_frame$Sex)

#Create an empty numeric matrix to fill with distances
SEXM<-array(0,c(nrow(Sex_frame),nrow(Sex_frame)))

#Derive matrix with binary sex similarity between each sample
for(i in 1:nrow(Sex_frame)){
  for(j in 1:nrow(Sex_frame)){ 
    if(Sex_frame$Sex[i]==Sex_frame$Sex[j]){
      SEXM[i,j]= 1
    } else{
      SEXM[i,j]= 0
    }
  }
} 

#Note that Sex similarity matrix has rownames and colnames in the same order as key
all(rownames(SEXM)==key$ID)

#Save matrix to ready matrices folder
saveRDS(SEXM,"ready_matrices/SEXM.rds")

```
1.5 Dyadic data frame

These matrices can be used in models as such (see MRQAP example below) or they can be "unraveled" into dyadic data frames, where each row is one pair and columns mark the various calculated similarities and distances between members of this pair. 

Let's first unravel the matrices into one dyadic data frame:

```{r}
#Read in any of the matrices if not in already:
JACM<-readRDS("ready_matrices/JACM.rds")
TEMPM<-readRDS("ready_matrices/TEMPM.rds")
SPATM<-readRDS("ready_matrices/SPATM.rds")
AM<-readRDS("ready_matrices/AM.rds")
SEXM<-readRDS("ready_matrices/SEXM.rds")



#First unravel the matrices into vectors of the lower quantile of the matrix
jac<-c(as.dist(JACM))
am<-c(as.dist(AM))
spa<-c(as.dist(SPATM))
temp<-c(as.dist(TEMPM))
sex<-c(as.dist(SEXM))

data.dyad<-data.frame(Social_proximity=am,Microbiome_similarity=jac,spatial_distance=spa,Temporal_distance=temp,sex_similarity=sex) 

# extracting Individual-combinations of the matrices
list<-expand.grid(key$ID,key$ID) 
# This created individual-to-same-individual pairs as well. Get rid of these:
list<-list[which(list$Var1!=list$Var2),] # this still has both quantiles in--> add 'unique' key 
list$key <- apply(list, 1, function(x)paste(sort(x), collapse='')) 
list<-subset(list, !duplicated(list$key)) 
# sanity check that the Individual name combinations are in the same order as the lower quantile value vector of the matrix
JACM[which(rownames(JACM)==list$Var1[3]),which(colnames(JACM)==list$Var2[3])]==jac[3]
# add the names of both individuals participating in each pair into the data
data.dyad$IDA<-list$Var2
data.dyad$IDB<-list$Var1

# Make sure you have got rid of all self comparisons
data.dyad<-data.dyad[which(data.dyad$IDA!=data.dyad$IDB),] 

```
2. MRQAP model

This is a matrix permutation model that randomizes the residuals from the regression on each predictor 1000 times and compares the observed correlation to correlations in the resulting null models.See Dekker et al (2007) and asnipe package vignette by Damien Farine for details
```{r}
library(asnipe)

#Scale values in all matrices between 0-1, unless they are already naturally between this range
#scaling function:
range.use <- function(x,min.use,max.use){ (x - min(x,na.rm=T))/(max(x,na.rm=T)-min(x,na.rm=T)) * (max.use - min.use) + min.use }

SPATM_scaled<-range.use(SPATM,0,1)
TEMPM_scaled<-range.use(TEMPM,0,1)

model1<- mrqap.dsp(JACM~ SEXM+SPATM_scaled+TEMPM_scaled+AM , directed="undirected", diagonal=FALSE, randomisations=1000)

model1
```
3. Dyadic Bayesian regression using MCMCglmm
```{r}
library(MCMCglmm)

#Scale values in all matrices between 0-1, unless they are already naturally between this range
#scaling function:
range.use <- function(x,min.use,max.use){ (x - min(x,na.rm=T))/(max(x,na.rm=T)-min(x,na.rm=T)) * (max.use - min.use) + min.use }

# variables to scale
scalecols<-c("spatial_distance","Temporal_distance")

#scaling:
for(i in 1:ncol(data.dyad[,which(colnames(data.dyad)%in%scalecols)])){
  data.dyad[,which(colnames(data.dyad)%in%scalecols)][,i]<-range.use(data.dyad[,which(colnames(data.dyad)%in%scalecols)][,i],0,1)
}

#Model
model2<-MCMCglmm(Microbiome_similarity~1+sex_similarity+spatial_distance+Temporal_distance+Social_proximity , 
             data=data.dyad,
             family= "gaussian",
             random =~ mm(IDA+IDB),
             verbose=FALSE)

summary(model2)

```         

I would actually recommend using models from brms package for this, as they allow betaregression more suitable for proportional response variables. But they are very slow so using MCMCglmm as a work horse here as the values of Jaccard are roughly normal and gaussian regression is generally OK.